Biology background

### 2014-10-30 Rmarkdown/biology_background.Rmd
Writer, reads and erasers of DNAm. Bisulfite-conversion

- [x] Done

@@

A common analogy used in describing DNA methylation is that of "writers", "readers" and "erasers" [@Moore:2013in]. Writers catalyse the methyl group onto the DNA, readers recognise methylated DNA and erasers remove the methyl group from the DNA.

In mammalian cells, the writers are the DNA _methyltransferase_ (DNMT) enzymes. These DNMTs are commonly split into two groups, namely the maintainence methyltransferase (thought to be DNMT1) and the _de novo_ methyltransferases (thought to be DNMT3a and DNMT3b)[^DNMT]. DNA methylation is not preserved by the DNA replication machinery and so it is the role of DNMT1 to restore the methylation pattern on the daughter strand of DNA following DNA replication. In contrast, DNMT3a and DNMT3b lay down new methylation marks and are particularly active during development when there are widespread changes in DNA methylation (__SOURCE__). DNMT1 and DNMT3b appear to be essential for mammalian development as mouse knockouts[^KO] for either gene is embryonically lethal (__SOURCE: http://labs.genetics.ucla.edu/fan/papers/npp2012112a.pdf__) whereas mouse knockouts for DNMT3a are runted but survive for ~4 weeks after birth (__SOURCE: http://labs.genetics.ucla.edu/fan/papers/npp2012112a.pdf__).

[^DNMT]: DNMT2, now known as TRDMT1, was once thought to be a DNA methyltransferase, however, __http://www.sciencemag.org/content/311/5759/395__ showed that it in fact methylates a small RNA and not DNA. Another protein, DNMT3L, is homologous to DNMT3a and DNMT3b but does contain catalytic domain that is necessary for methyltransferase activity. Instead, DNMT3L is thought to stimulate the activity of DNMT3a and DNMT3b (__SOURCE__).

@@

The readers of DNA methylation recognise methylated DNA. These readers can recruit additional proteins to the site of the methylated cytosine to perform a variety of functions related to gene expression. For example, the _MBD_ (methyl-CpG-binding domain) group of proteins bind to DNA containing a methylated CpG which then suppress gene expression by preventing transcription factor binding at that site (__SOURCE__). Another group, the _UHRF_ (ubiquitin-like, containing PHD and RING finger domain) proteins, help DNMT1 methylate _hemimethylated_ DNA, which is DNA where one strand is methylated and the other is not (e.g. the parent strand compared to the daughter strand following DNA replication).

The removal or erasure of DNA methylation, _demethylation_, is characterised as _passive_ loss or _active_ removal. Passive loss occurs when the maintainence methyltransferases do not efficiently perform their role of restoring DNA methylation following cell division. This leads to a gradual, stochastic and genome-wide loss of DNA methylation after multiple cell divisions. This form of passive demethylation, sometimes called replication-dependent demethylation, cannot explain observations of precise targeted demethylation (__EXAMPLE__) nor the two stages of rapid global demethylation that occur during development [@Wu:2014gw].

Active demethylation is currently a hot topic in epigenetics research. Multiple mechanisms have been proposed, and it is indeed likely that there are multiple ways to achieve active demethylation. These mechanisms were recently reviewed by [@Wu:2014gw], which I now briefly summarise:

1. The direct removal of the methyl group from 5mC is considered unlikely due to the strong carbon-carbon bond between the methyl group and the cytosine.
2. There is evidence that the DNA repair machinery can be co-opted to remove a methylated base or the surrounding region. The excised base or region is then repaired with unmethylated cytosines replacing 5mCs.
3. 5mC oxidation-dependent active DNA demethylation. This follows from the observation that the TET enzymes can iteratively oxidate a 5mC -> 5hmC -> 5fC -> 5caC reaction. The removal of 5hmC, 5fC or 5caC is biochemically "easier" than the removal of 5mC and could occur via a more efficient form of replication-dependent demethylation, direct removal of the oxidized methyl group or through the DNA repair machinery (__FIGURE OF TET-CYCLE__).

One question raised by point (3) is whether 5hmC, 5fC and 5caC are simply intermediate products in an active demethylation cycle or if they themselves are bona fide epigenetic marks.

@@

- [ ] Done

@@

A challenge to measuring DNA methylation is that it is erased by standard molecular biology techniques, such as _polymerase chain reaction_ (PCR) and bacterial cloning, and it is not revealed by DNA hybridization assays [@Laird:2010iv]. Therefore, almost all assays of DNA methylation require one of the following _pre-treatments_ of the DNA:

@@

An exception to this classification scheme are a new class of assay that seek to directly "read" whether a position is methylated or unmethylated without requiring a pre-treatment of the DNA. For example, both [@Laszlo:2013kf, Schreiber:2013in] measure the change in current as a DNA molecule passes through a nanopore to infer whether a cytosine was methylated while __PACBIO AND ONT DO WHAT?__.

Almost all assays of DNA methylation measure a "population average" from a pool of hundreds or thousands of cells. For a diploid organism, this is an average over several different levels: the two DNA strands, the two homologous chromosomes within a diploid cell and the hundreds or thousand of cells used in the assay (see __FIGURE__). Hundreds or thousands of cells are required in order to have sufficient material as input for the assay. Assays that require only a single cell as input do exist but are still in development and not in widespread use (__SOURCE/EXAMPLES__).

@@

[^resolution]: Depending on the experiment and its aims, the resolution of an assay might instead be defined as the scale on which DNA methylation can be quantified or that which allows inference to address a specific hypothesis.

The throughput of an assay can be quantified in two ways. The first is per-sample throughput, which is how many measurements of DNA methylation are made per-sample[^throughput1]. This is typically what people mean when they describe an assay as being "high-throughput" or "low-throughput" and is the definition I use in the title of my thesis. Depending on your definition of "high", a high-througput assay will produce on the order of tens of thousands to billions of measurements per sample. The second definition of throughput is per-dollar, that is, "how many samples can I process within my budget?". To avoid confusion, __I should refer to the latter definition as something else__.

@@

The choice of which assay to use for an experiment is a trade-off between resolution, per-sample throughput and per-dollar throughput. Experiments that use an assay with high-resolution and high per-sample throughput generally have fewer samples than experiments using a lower resolution assay or an assay with lower per-sample throughput.

@@

- [ ] Done

_Restriction endonucleases_ are an important technique in molecular biology. These enzymes can preferrentially "cut/cleave/digest" DNA at particular sequence motifs. The motif at which a restriction enzyme cleaves DNA is called the _recognition motif_ or _restriction sequence_. The methylation of a position in the recognication motif can inhibit a restriction enzyme from cleaving the DNA. This can be used to design an assay to infer the methylation state of a DNA fragment.

For example, the recognition site of the restriction enzyme _HpaII_ is `CCGC`. However, _HpaII_ will only digest DNA when the second cytosine in the motif is unmethylated. The _HELP_ (_HpaII_ tiny fragment enrichment by ligation-mediated PCR) assay compares DNA digested by _HpaII_ to one digested with another restriction enzyme that has the same recognition motif but is methylation-insensitive (_MspI_) to identify _hypomethylated_ regions of a genome (__SOURCE__).

Another popular restriction enzyme for studying DNA methylation is _McrBC_, "an enzyme with the unusual and desirable property of cutting methylated DNA promiscuously (recognition sequence $R^{m} C(N)_{55– 103} R^{m} C$)" [@Irizarry:2008hg].

Assays based on restriction enzyme were some of the first developed for studying DNA methylation. They were initially developed for studying particular loci although they have been extended to genome-scale analysis approaches [@Laird:2010iv]. Restriction enzyme assays have a relatively low resolution (__REASONS/SOURCE__) but may also be used as a first-pass to enrich a sample for methylated or unmethylated regions which are then assayed using a higher resolution technique.

@@

- [ ] Done

Affinity enrichment assays compare measurements between an "enriched" version and an "input" (control) version of the same sample to infer the presence or absence of DNA methylation. This is very similar to the idea behind chromatin immunoprecipitation followed by microarray hybridisation (ChIP-chip) or sequencing (ChIP-seq). Enrichment may be done using a methylation-specific antibody, such as __EXAMPLE__, or a __unmethylation-specific(?)__ antibody, such as __EXAMPLE__.

Some examples of affinity enrichment assays for DNA methylation are the microarray-based MeDIP, mDIP and mCIP (all based on methylated DNA immunoprecipitation) and their sequencing-based relatives, MeDIP-seq and mDIP-seq. These are all low resolution assays since they are based on the enrichment of regional differences between the enriched and input samples. Furthermore, the bioinformatic analysis of data from these assays is complicated by the varying CpG density along the genome, which leads to different enrichment affinities for different regions of the genome. However, these assays are can provide a relatively cheap and efficient genome-wide assessment of DNA methylation [@Laird:2010iv}].

@@

- [ ] Done

In the 1980s, __Hayatsu OR Wang et al.__ discovered that when _denatured_ (__DEFINE?__) DNA is treated with sodium bisulfite (__CHEMICAL SYMBOL__), unmethylated cytosines deaminate to uracils much faster than do methylated cytosines (__SOURCE__). The methylated cytosines are said to be "protected" from conversion to uracils. When the bisulfite treated is amplified by PCR the uracils are subsequently converted to thymines.

This discovery led to the development of assays for studying studying cytosine methylation based on the pre-treatment of DNA with sodium bisulfite, which are sometimes referred to as bisulfite-conversion assays (__FROMMER; CLARK__). These assays are all based on the idea of comparing the sequence of the untreated DNA to the sequence of the bisulfite treated DNA to infer the methylation state of all cytosines in the sequence by whether or not they were converted to uracil (thymine) following the bisulfite treatment (__see FIGURE__).

@@

- [ ] Done

>>>>

@@

A common analogy used in describing DNA methylation uses 'writers', 'readers' and 'erasers' \cite{Moore:2013in}. Writers catalyse the methyl group onto the DNA, readers recognise methylated DNA and erasers remove the methyl group from the DNA.

In mammalian cells, the writers are the DNA _methyltransferase_ (DNMT) enzymes. These DNMTs are commonly split into two groups, namely the maintainence methyltransferase (thought to be DNMT1) and the _de novo_ methyltransferases (thought to be DNMT3a and DNMT3b)[^DNMT]. DNA methylation is not preserved by the DNA replication machinery and so it is the role of DNMT1 to restore the methylation pattern on the daughter strand of DNA following DNA replication. In contrast, DNMT3a and DNMT3b lay down new methylation marks and are particularly active during development when there are widespread changes in DNA methylation (__TODO: cite__). DNMT1 and DNMT3b appear to be essential for mammalian development as mouse knockouts[^KO] for either gene is embryonically lethal \cite{Li:1992bp} whereas mouse knockouts for DNMT3a are runted but survive for ~4 weeks after birth \cite{Li:1992bp}.

[^DNMT]: DNMT2, now known as TRDMT1, was once thought to be a DNA methyltransferase, however, it was shown to in fact methylate a small RNA and not DNA \cite{Goll:2006kb}. Another protein, DNMT3L, is homologous to DNMT3a and DNMT3b but does contain catalytic domain that is necessary for methyltransferase activity. Instead, DNMT3L is thought to stimulate the activity of DNMT3a and DNMT3b (__TODO: cite__).

@@

The readers of DNA methylation recognise methylated DNA. These readers can recruit additional proteins to the site of the methylated cytosine to perform a variety of functions related to gene expression. For example, the methyl-CpG-binding domain (_MBD_) group of proteins bind to DNA containing a methylated CpG, which can then suppress gene expression by preventing transcription factor binding at that site \cite{Nan:1993gm}. Another group, the ubiquitin-like, containing PHD and RING finger domain (_UHRF_) proteins, help _DNMT1_ methylate hemimethylated DNA, such as the daughter strand created during DNA replication \cite{Sharif:2007ft, Bostick:2007jh}.

The removal or erasure of DNA methylation, _demethylation_, is characterised as _passive_ loss or _active_ removal. Passive loss occurs when the maintenance methyltransferases do not efficiently perform their role of restoring DNA methylation following cell division. This leads to a gradual, stochastic and genome-wide loss of DNA methylation after multiple cell divisions. This form of passive demethylation, sometimes called replication-dependent demethylation, cannot explain observations of precise targeted demethylation (__TODO: example__) nor the two stages of rapid global demethylation that occur during development \cite{Wu:2014gw}.

Active demethylation is currently a hot topic in epigenetics research. Multiple mechanisms have been proposed, and it is indeed likely that there are multiple ways to achieve active demethylation. These mechanisms were recently reviewed by \cite{Wu:2014gw}, which I briefly summarise:

1. The direct removal of the methyl group from _5mC_ is considered unlikely due to the strong carbon-carbon bond between the methyl group and the cytosine.
2. There is evidence that the DNA repair machinery can be co-opted to remove a methylated base or the surrounding region. The excised base or region is then repaired with unmethylated cytosines replacing 5mCs.
3. 5mC oxidation-dependent active DNA demethylation. This follows from the observation that the TET enzymes can iteratively oxidate a _5mC_ -> _5hmC_ -> _5fC_ -> _5caC_ reaction. The removal of _5hmC_, _5fC_ or _5caC_ is biochemically "easier" than the removal of _5mC_ and could occur via a more efficient form of replication-dependent demethylation, direct removal of the oxidized methyl group or through the DNA repair machinery (__TODO: Figure of TET cycle__).

One question raised by point (3) is whether _5hmC_, _5fC_ and _5caC_ are simply intermediate products in an active demethylation cycle or if they themselves are bona fide epigenetic marks.

@@

@@

A challenge to measuring DNA methylation is that it is erased by standard molecular biology techniques, such as the _polymerase chain reaction_ (PCR) and bacterial cloning, and it is not revealed by DNA hybridization assays \cite{Laird:2010iv}. Therefore, almost all assays of DNA methylation require one of the following _pre-treatments_ of the DNA:

@@

An exception to this classification scheme are a new class of assay that seek to directly "read" whether a position is methylated or unmethylated without requiring a pre-treatment of the DNA. For example, both \cite{Laszlo:2013kf} and \cite{Schreiber:2013in} measure the change in current as a DNA molecule passes through a nanopore to infer whether a cytosine was methylated. Similarly, __TODO: what does Pacbio and Oxford Nanopore do?__

Almost all assays of DNA methylation measure a "population average" from a pool of hundreds or thousands of cells. For a diploid organism, this is an average over several different levels: the two DNA strands, the two homologous chromosomes within a diploid cell and the hundreds or thousand of cells used in the assay (see __FIGURE__). Hundreds or thousands of cells are required in order to have sufficient material as input for the assay. Assays that require only a single cell as input do exist but are still in development and not in widespread use (__TODO: sources and examles__).

@@

[^resolution]: Depending on the experiment and its aims, the resolution of an assay might instead be defined as the scale on which DNA methylation can be quantified or the scale at which allows inference to address a specific hypothesis.

The throughput of an assay can be quantified in two ways. The first is per-sample throughput, which is how many measurements of DNA methylation are made per-sample[^throughput1]. This is typically what people mean when they describe an assay as being "high-throughput" or "low-throughput" and is the definition I use in the title of my thesis. Depending on your definition of "high", a high-througput assay will produce on the order of tens of thousands to billions of measurements per sample. The second definition of throughput is  related to cost, be it money or time, i.e., "how many samples can I process?".

@@

The choice of which assay to use for an experiment is a trade-off between resolution, per-sample throughput and per-cost throughput. Experiments that use an assay with high-resolution and high per-sample throughput generally have fewer samples than experiments using a lower resolution assay or an assay with lower per-sample throughput.

@@

_Restriction endonucleases_ are an important technique in molecular biology. These enzymes can preferentially "cut/cleave/digest" DNA at particular sequence motifs. The motif at which a restriction enzyme cleaves DNA is called the _recognition motif_ or _restriction sequence_. The methylation of a position in the recognition motif can inhibit a restriction enzyme from cleaving the DNA. This can be used to design an assay to infer the methylation state of a DNA fragment.

For example, the recognition site of the restriction enzyme _HpaII_ is `CCGC`. However, _HpaII_ will only digest DNA when the second cytosine in the motif is unmethylated. The _HELP_ (_HpaII_ tiny fragment enrichment by ligation-mediated PCR) assay compares DNA digested by _HpaII_ to one digested with another restriction enzyme that has the same recognition motif but is methylation-insensitive (_MspI_) to identify _hypomethylated_ regions of a genome \cite{Khulan:2006gx}.

Assays based on restriction enzyme were some of the first developed for studying DNA methylation. They were initially developed for studying particular loci although they have been extended to genome-scale analysis approaches \cite{Laird:2010iv}. Restriction enzyme assays have a relatively low resolution (__TODO: reasons and source__) but may also be used as a first-pass to enrich a sample for methylated or unmethylated regions which are then assayed using a higher resolution technique.

@@

Affinity enrichment assays compare measurements between an 'enriched' version and an 'input' (control) version of the same sample to infer the presence or absence of DNA methylation. This may use antibody immunoprecipitation or methyl-binding proteins. Some examples of affinity enrichment assays for DNA methylation are the microarray-based MeDIP, mDIP and mCIP and their sequencing-based relatives, MeDIP-seq and mDIP-seq.

These are all low resolution assays since they are based on the enrichment of regional differences between the enriched and input samples. Furthermore, the bioinformatic analysis of data from these assays is complicated by the varying CpG density along the genome, which leads to different enrichment affinities for different regions of the genome. However, these assays are can provide a relatively cheap and efficient genome-wide assessment of DNA methylation \cite{Laird:2010iv}.

@@

In the 1980s, two research groups independently discovered that when DNA is treated with sodium bisulfite ($NaHSO_{3}$), unmethylated cytosines deaminate to uracils much faster than do methylated cytosines \cite{Shapiro:1970im, Hayatsu:1970jv}. The methylated cytosines are said to be "protected" from conversion to uracils. This discovery led to the development of assays for studying studying cytosine methylation based on the pre-treatment of DNA with sodium bisulfite \cite{Frommer:1992wq}, which are referred to as bisulfite-conversion assays.

When bisulfite-treated is amplified by PCR, the uracils are converted to thymines. Thus, these assays are all based on the idea of comparing the sequence of the untreated DNA to the sequence of the bisulfite treated DNA to infer the methylation state of all cytosines in the sequence by whether or not they were converted to uracil/thymine following the bisulfite treatment (__FIGURE__).

@@

__UP TO HERE__
